Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-11-12 WSSSPE3/report/WSSSPE3_review.tex
answering questions, fixing some text

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}} Lorraine Hwang presented experiences with CIG: the Computational
Infrastructure for Geodynamics, a community software with a worldwide user base.
Like others, their main goals include high usability, sustainability, and
reproducibility. In order to achieve these goals, communication channels have been
developed, such as mailing lists, wikis, workshops, hackathons, tutorials, and
webinars. In order to contribute to the infrastructure, codes have to adhere to
specified minimum standards, but ideals \choinote{ideals or ideas?} are also given to encourage reaching the
aforementioned goals. These include, e.g., the use of version control, certain
coding styles, the presence and nature of code tests and documentation, and
certain user workflows.

@@

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the HEP \choinote{What is HEP?} effort to Biology,
Astrophysics, Digital Curation, and other disciplines. It includes physicists,
digital librarians, as well as computer scientists, and aims to achieve some
commonality across disciplines. Examples are meta-data descriptions of archived
data, computational descriptions, descriptions of how data was processed, 
questions such as whether computation replication can be
automated, and what the impact of access policies on the preservation
infrastructure is. One of the products of this effort is a suite of tools that
deals with this preservation, and the questions was posed whether that software
itself is sustainable. Points that were brought up included the need for a user
community depending on a given software, and the need to provide added value for
its users. An important method to achieve this was to work with the user
community from the start, and to budget that way. For the specific example of
the preservation software, this means that besides adding value to the
community, it needs to be transparent to their workflows, i.e., not requiring
additional effort to preserve ``research objects''.

@@

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}} Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, great developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design and development and testing phase \choinote{very complicated phrase of the phase}. The main
deficiency with this approach was pointed out to be the too-late
interaction with users. Instead, an alternative approach was proposed that, at
first, skips the development phase and repeatedly goes through requirement
gathering, design and user interaction/learning phases, and only eventually
starts development once an agreement is reached, leading to an increased user
satisfaction, higher adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and Sergei
Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
Biology: Lessons Learned on the KBase Project}} In this talk, Dan Gunter
presented experiences and lessons learned as part of KBase, an open software and
data platform for addressing the grand challenge of systems biology: predicting
and designing biological function. KBase is a unified system that integrates
data and analytical tools for comparative functional genomics of microbes,
plants, and their communities. However, it is also a collaborative environment
for sharing methods and results, and placing those results in the context of
knowledge in the field. Being a large, multi-institutional project, one of the
big challenges is to agree on standard to enable a single, maintainable system.
Working in isolation does not work (anymore) within this field, and they \choinote{who are they?} also
didn't have standards for software engineering. This is contrasted to computer
science research, where these approaches shorten design cycles, leading to more
time for highly rewarded activities like publishing, performance studies,
graduating students, or protecting ideas before publication. Sustainability is
not of big concern there, compared to these. \katznote{prev sentence is unclear}
Instead of this more traditional approach, KBase uses the following for large,
distributed teams. After picking projects and team members, 4-5 teams work on
projects about two weeks before a 1-week evaluation by the executive committee,
and after which new teams and projects might be chosen to restart the ``agile''
development cycle. This process is intentionally open and documented.

>>>>

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}} Lorraine Hwang presented experiences with CIG: the Computational
Infrastructure for Geodynamics, a community software with a worldwide user base.
Like others, their main goals include high usability, sustainability, and
reproducibility. In order to achieve these goals, communication channels have been
developed, such as mailing lists, wikis, workshops, hackathons, tutorials, and
webinars. In order to contribute to the infrastructure, codes have to adhere to
specified minimum standards, but ideals %\choinote{ideals or ideas?} \katznote{ideals}
are also given to encourage reaching the
aforementioned goals. These include, e.g., the use of version control, certain
coding styles, the presence and nature of code tests and documentation, and
certain user workflows.

@@

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the high energy physics effort to biology,
astrophysics, digital curation, and other disciplines. It includes physicists,
digital librarians, as well as computer scientists, and aims to achieve some
commonality across disciplines. Examples are meta-data descriptions of archived
data, computational descriptions, descriptions of how data was processed, 
questions such as whether computation replication can be
automated, and what the impact of access policies on the preservation
infrastructure is. One of the products of this effort is a suite of tools that
deals with this preservation, and the questions was posed whether that software
itself is sustainable. Points that were brought up included the need for a user
community depending on a given software, and the need to provide added value for
its users. An important method to achieve this was to work with the user
community from the start, and to budget that way. For the specific example of
the preservation software, this means that besides adding value to the
community, it needs to be transparent to their workflows, i.e., not requiring
additional effort to preserve ``research objects''.

@@

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}} Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, great developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design, development, and testing phase. The main
deficiency with this approach was pointed out to be the too-late
interaction with users. Instead, an alternative approach was proposed that, at
first, skips the development phase and repeatedly goes through requirement
gathering, design and user interaction/learning phases, and only eventually
starts development once an agreement is reached, leading to an increased user
satisfaction, higher adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and Sergei
Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
Biology: Lessons Learned on the KBase Project}} In this talk, Dan Gunter
presented experiences and lessons learned as part of KBase, an open software and
data platform for addressing the grand challenge of systems biology: predicting
and designing biological function. KBase is a unified system that integrates
data and analytical tools for comparative functional genomics of microbes,
plants, and their communities. However, it is also a collaborative environment
for sharing methods and results, and placing those results in the context of
knowledge in the field. Being a large, multi-institutional project, one of the
big challenges is to agree on standard to enable a single, maintainable system.
Working in isolation does not work (anymore) within this field, and the community
in the field also
didn't have standards for software engineering. This is contrasted to computer
science research, where these approaches shorten design cycles, leading to more
time for highly rewarded activities like publishing, performance studies,
graduating students, or protecting ideas before publication. Sustainability is
not of big concern there, compared to these. \katznote{prev sentence is unclear}
Instead of this more traditional approach, KBase uses the following for large,
distributed teams. After picking projects and team members, 4-5 teams work on
projects about two weeks before a 1-week evaluation by the executive committee,
and after which new teams and projects might be chosen to restart the ``agile''
development cycle. This process is intentionally open and documented.
